Iterative Monomorphisation

### 2025-06-26 iterative_monomorphisation.tex
edits

To detect if overfitting took place, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems. We carried out the rest of the evaluations on the remaining 1034 problems.

Before finding values to assign to the bounds of the monomorphisation algorithm, we must choose which base options to run Zipperposition with. The prover has a portfolio mode with several configurations. Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.

@@

\begin{itemize}
   \item \PolyCap: \(\infty\)
   \item \PolyMult: \(1\)
   \item \PolyFloor: \(50\)
   \item substitution ordering: separation
   \item \SubstLimit: \(\infty\)
   \item \MonoSubstsLimit: \(10\)
\end{itemize}

@@

Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and per iteration. The limit on newly generated type arguments is determined by three components that form a natural group of bounds. The table shows that generating no new monomorphic type arguments seems to be the best approach. This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments. This is because non-monomorphic type argument generation can generate substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.

@@

%The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
The substitution generation phase occurs once all type arguments have been generated. The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated. The %substitution ordering
heuristic greatly affects monomorphising substitution generation. The `age' ordering of substitution orders substitutions generated in earlier iterations first. The `random' ordering randomly shuffles substitutions. Finally, the `separation' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seem to affect performance when the `separation' heuristic is used.
% TODO find a better term than "separation heuristic" maybe separation ?

@@

\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \multicolumn{3}{c}{substitution ordering} \\
   \multirow{1}{6em}{mono subst} & age & random & separation\\
   \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
   2  & 161 & 178 & 175 \\
   5  & 161 & 178 & 180 \\
   7  & 161 & 178 & 182 \\
   10 & 161 & 178 &\bf{184} \\
   \bottomrule
\end{tabular}
\label{subst_gen_table}
\end{table}

@@

For larger problems, the monomorphisation algorithm may time out despite the bounds. In these cases neither Zipperposition nor E will have had a chance to try to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iterations of the monomorphisation algorithm. Neither parameter seems to substantially affect the algorithm's performance.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. Unfortunately we needed to exclude Vampire because of parsing issues.
%Vampire failed to parse a significant number of the benchmarks problems.
The fix was unavailable for Vampire's higher order branch at the time of the evaluation.

@@

We described an algorithm for iteratively instantiating polymorphic types to
produce monomorphic problems as a preprocessor. Our primary motivation was to
improve the success rate of Zipperposition and its monomorphic E backend, and
indeed our evaluation shows a clear improvement. We also saw that even with
automatic provers that support polymorphism, iterative monomorphisation is a
better alternative in practice.

>>>>

To guard against overfitting took place, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems. We carried out the rest of the evaluations on the remaining 1034 problems.

Before finding values to assign to the bounds of the monomorphisation algorithm, we must choose which base options to run Zipperposition with. The prover has a portfolio mode with several configurations. Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
The preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.

@@

\begin{itemize}
   \item \PolyCap: \(\infty\)
   \item \PolyMult: \(1\)
   \item \PolyFloor: \(50\)
   \item substitution order: separation
   \item \SubstLimit: \(\infty\)
   \item \MonoSubstsLimit: \(10\)
\end{itemize}

@@

Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and per iteration. The limit on newly generated type arguments is determined by three components that form a natural group of bounds.

The table shows that generating no new monomorphic type arguments seems to be the best approach. This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments. This is because non-monomorphic type argument generation can produce substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.

@@

%The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution order heuristic has a significant impact on monomorphising substitution generation. The ``age'' order of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' order randomly shuffles substitutions. Finally, the ``separation'' order separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the order heuristic.
The substitution generation phase occurs once all type arguments have been generated. The bound limiting the number of monomorphising substitutions per formula is directly related to the order which dictates how such monomorphising substitutions are generated. The %substitution order
heuristic greatly affects monomorphising substitution generation. The `age' order of substitution orders substitutions generated in earlier iterations first. The `random' order randomly shuffles substitutions. Finally, the `separation' order separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions seem to affect performance only when the `separation' heuristic is used.
% TODO find a better term than "separation heuristic" maybe separation ?

@@

\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \multicolumn{3}{c}{substitution order} \\
   \multirow{1}{6em}{mono subst} & age & random & separation\\
   \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
   2  & 161 & 178 & 175 \\
   5  & 161 & 178 & 180 \\
   7  & 161 & 178 & 182 \\
   10 & 161 & 178 &\bf{184} \\
   \bottomrule
\end{tabular}
\label{subst_gen_table}
\end{table}

@@

For larger problems, the monomorphisation algorithm may time out despite the bounds. In these cases, neither Zipperposition nor E will have had a chance to try to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iterations of the monomorphisation algorithm. Neither parameter seems to substantially affect the algorithm's performance.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. Unfortunately we needed to exclude Vampire because of parsing issues.
%Vampire failed to parse a significant number of the benchmarks problems.
The fix for these issues was unavailable for Vampire's higher order branch at the time of the evaluation.

@@

We described a translation algorithm that iteratively instantiates polymorphic types to
produce monomorphic problems. Our primary motivation was to improve the success
rate of Zipperposition and its monomorphic E backend, and indeed our evaluation
shows a clear improvement. We also saw that even with automatic provers that
support polymorphism, iterative monomorphisation is a better alternative in
practice.
